Baseline Drift Estimation for Air Quality Data Using Quantile Trend Filtering

### 2019-03-30 Manuscript/detrendify_aoas.tex
new threshold choice

In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. To reduce cost and power consumption of the SPod, the temperature and relative humidity of the air presented to the photoionization detectors (PIDs) was not controlled and as a result the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, which is illustrated by the three sharp transient spikes at 11:32, 14:10, and 16:03. However, the baseline drift varies from one sensor to another, obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor %using $\ell_1$-quantile-trend filtering
and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed vector $\V{y}$ into a slowly varying component $\V{\theta}$ and a transient component $\V{s}$ can lead to greatly improved and simplified downstream analysis.

@@

%% ----------------------------------------------------------------------
%% Computational Complexity and Convergence
%% ----------------------------------------------------------------------
\subsection{Computational Complexity and Convergence}

@@

An important practical issue in baseline estimation is the choice of regularization parameter $\lambda$, which controls the degree of smoothness in $\V{\theta}$. In this section, we introduce four methods for choosing $\lambda$. The first is a validation based approach; the latter three are based on information criteria.

@@

\begin{eqnarray}
\label{eq:validate}
&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\tilde{\rho}_{\tau_j}(\V{y} - \V{\theta}_{j}) +
\lambda \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ],
\end{eqnarray}
which can be solved via \Alg{admm} with trivial modification to the quadratic program sub-problems. We fit a sequence of values for $\lambda$ setting $\lambda_j = \lambda$ for all $j$.  Then for each quantile level $j$ we select the $\lambda_j$ that minimizes the hold-out prediction error quantified by $\breve{\rho}_{\tau_j}(\V{y} - \Vhat{\theta}_j(\lambda))$ where $\Vhat{\theta}_j(\lambda)$ is the solution to \Eqn{validate} and
\begin{eqnarray*}
	\breve{\rho}_{\tau}(\V{r}) & = & \sum_{i \in \mathcal{V}} \VE{r}{i}(\tau-\One(\VE{r}{i}<0)).
\end{eqnarray*}

\subsection{Information Criteria}
\cite{KoenkerNgPortnoy1994} addressed the choice of regularization parameter by proposing the Schwarz criterion for the selection of $\lambda$
\begin{eqnarray*}
	\label{eq:SIC}
	\mbox{SIC}(p_{\lambda}) & = & \log\left[\frac{1}{n}\rho_{\tau}(\V{y} - \V{\theta})\right] + \frac{1}{2n}p_{\lambda}\log n.
\end{eqnarray*}
where $p_{\lambda} = \sum_i \One(\VE{y}{i} = \widehat{\theta}_i)$ is the number of interpolated points, which can be thought of as active knots. Equivalently, $p_{\lambda}$ can be substituted with the number of non-zero components in $\M{D}^{(k+1)}\V{\theta}$ which we denote $\nu$ and have found to be more numerically stable.  The SIC is based on the traditional Bayesian Information Criterion (BIC) which is given by
\begin{equation}
\mbox{BIC}(\nu) = -2\log(L\{\Vhat{\theta}\}) + \nu\log n
\end{equation}
where $L$ is the likelihood function. If we take the approach used in Bayesian quantile regression \citep{yu2001bayesian}, and view minimizing the check function as maximizing the asymmetric Laplace likelihood,

\begin{eqnarray*}
	L(\V{y} \mid \V{\theta}) & = & \left[\frac{\tau^n(1-\tau)}{\sigma}\right]^n\exp\left\{-\sum_i\rho_\tau\left(\frac{\VE{y}{i} - \VE{\theta}{i}}{\sigma}\right)\right\},
\end{eqnarray*}
we can compute the BIC as
\begin{eqnarray*}
	\mbox{BIC}(\nu) = 2\frac{1}{\sigma}\rho_{\tau}(\V{y}-\Vhat{\theta}) + \nu\log n
\end{eqnarray*}
where $\Vhat{\theta}$ is the estimated trend, and $\nu$ is the number of non-zero elements of $\Mn{D}{k+1}\Vhat{\theta}$. We can choose any $\sigma>0$ and have found empirically that $\sigma =  \frac{1-|1-2\tau|}{2}$ produces stable estimates.

@@

where $P$ is the total number of possible parameters and $\nu$ is the number of non-zero parameters included in given mod. We used this criteria with $\gamma = 1$, and $P=n-k-1$.  
As in the validation method, to chose $\V{\lambda}$ using the information criteria we fit a sequence of values for $\lambda$ setting $\lambda_j = \lambda$ for all $j$.  For each quantile level $j$, we select the $\lambda_j$ that minimizes the SIC, BIC, or eBIC value evaluated using $\tau_j$ and $\V{\theta}_j$. In the simulation study, we compare the performance of the SIC, scaled eBIC, and validation methods.

@@

\label{sec:application}
The low-cost ``SPod" air quality sensors output a time series that includes a slowly varying baseline,
the sensor response to pollutants, and high frequency random noise. These sensors record measurements every second are used to monitor pollutant concentrations at the fence lines of industrial facilities and detect time points when high concentrations are present. Ideally, three co-located and time aligned sensors (as shown in \Fig{raw_spod}) responding to a pollutant plume would result in the same signal classification after baseline trend removal and proper threshold choice. We first illustrate the difference between our \texttt{detrend\_eBIC} and the \texttt{qsreg} using just 2 hours of measurements so that the methods can be easily visualized and the \texttt{qsreg} method can be fit to the entire dataset \Sec{short-app}. We then compare the methods on an entire week of measurements, fitting the \texttt{qsreg} method to non-overlapping windows of the data \Sec{long-app}.

>>>>

In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. Due to changes in temperature and relative humidity the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, which is illustrated by the three sharp transient spikes at 11:32, 14:10, and 16:03. However, the baseline drift varies from one sensor to another, obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor %using $\ell_1$-quantile-trend filtering
and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed vector $\V{y}$ into a slowly varying component $\V{\theta}$ and a transient component $\V{s}$ can lead to greatly improved and simplified downstream analysis.

@@

@@

An important practical issue in baseline estimation is the choice of the regularization parameter $\lambda$, which controls the degree of smoothness in $\V{\theta}$. In this section, we introduce four methods for choosing $\lambda$. The first is a validation based approach; the latter three are based on information criteria. Each of the criteria we compare is calculated for a single quantile ($\tau_j$). Rather than combine results over quantiles, we allow the value of $\lambda$ to vary by quantile resulting in $\V{\lambda} = \{\lambda_1, ..., \lambda_J\}$. To choose the best value for each $\lambda_j$, we first estimate all of the quantile trends using $\V{\lambda} = \lambda\mathbf{1}$ over a grid of values for $\lambda$. We then determine the $\lambda_j$ that maximizes the criteria chosen evaluated using $\tau_j$. Finally, we re-estimate the non-crossing trends with the optimal values for $\lambda_j$. A more thorough approach would involve fitting the model on a J dimensional grid of values for $\V{\lambda}$ but this is computationally infeasible.

@@

\begin{eqnarray}
\label{eq:validate}
&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\tilde{\rho}_{\tau_j}(\V{y} - \V{\theta}_{j}) +
\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ],
\end{eqnarray}
which can be solved via \Alg{admm} with trivial modification to the quadratic program sub-problems. For each quantile level $j$, we select the $\lambda_j$ that minimizes the hold-out prediction error quantified by $\breve{\rho}_{\tau_j}(\V{y} - \Vhat{\theta}_j(\lambda_j))$ where $\Vhat{\theta}_j(\lambda_j)$ is the solution to \Eqn{validate} and
\begin{eqnarray*}
	\breve{\rho}_{\tau}(\V{r}) & = & \sum_{i \in \mathcal{V}} \VE{r}{i}(\tau-\One(\VE{r}{i}<0)).
\end{eqnarray*}

\subsection{Information Criteria}
\cite{KoenkerNgPortnoy1994} addressed the choice of regularization parameter by proposing the Schwarz criterion for the selection of $\lambda$
\begin{eqnarray*}
	\label{eq:SIC}
	\mbox{SIC}(p_{\lambda}, \tau_j) & = & \log\left[\frac{1}{n}\rho_{\tau_j}(\V{y} - \V{\theta}_j)\right] + \frac{1}{2n}p_{\lambda}\log n.
\end{eqnarray*}
where $p_{\lambda} = \sum_i \One(\VE{y}{i} = \widehat{\theta}_i)$ is the number of interpolated points, which can be thought of as active knots. Equivalently, $p_{\lambda}$ can be substituted with the number of non-zero components in $\M{D}^{(k+1)}\Vhat{\theta}_j$ which we denote $\nu$ and have found to be more numerically stable.  The SIC is based on the traditional Bayesian Information Criterion (BIC) which is given by
\begin{equation}
\mbox{BIC}(\nu) = -2\log(L\{\Vhat{\theta}\}) + \nu\log n
\end{equation}
where $L$ is the likelihood function. If we take the approach used in Bayesian quantile regression \citep{yu2001bayesian}, and view minimizing the check function as maximizing the asymmetric Laplace likelihood,

\begin{eqnarray*}
	L(\V{y} \mid \V{\theta}) & = & \left[\frac{\tau^n(1-\tau)}{\sigma}\right]^n\exp\left\{-\sum_i\rho_\tau\left(\frac{\VE{y}{i} - \VE{\theta}{i}}{\sigma}\right)\right\},
\end{eqnarray*}
we can compute the BIC as
\begin{eqnarray*}
	\mbox{BIC}(\nu, \tau_j) = 2\frac{1}{\sigma}\rho_{\tau_j}(\V{y}-\Vhat{\theta}_j) + \nu\log n
\end{eqnarray*}
where $\Vhat{\theta}$ is the estimated trend, and $\nu$ is the number of non-zero elements of $\Mn{D}{k+1}\Vhat{\theta}$. We can choose any $\sigma>0$ and have found empirically that $\sigma =  \frac{1-|1-2\tau|}{2}$ produces stable estimates.

@@

where $P$ is the total number of possible parameters and $\nu$ is the number of non-zero parameters included in given mod. We used this criteria with $\gamma = 1$, and $P=n-k-1$. In the simulation study, we compare the performance of the SIC, scaled eBIC, and validation methods.

@@

\label{sec:application}
The low-cost ``SPod" air quality sensors output a time series that includes a slowly varying baseline,
the sensor response to pollutants, and high frequency random noise. These sensors record measurements every second are used to monitor pollutant concentrations at the fence lines of industrial facilities and detect time points when high concentrations are present. Ideally, three co-located and time aligned sensors (as shown in \Fig{raw_spod}) responding to a pollutant plume would result in the same signal classification after baseline trend removal and proper threshold choice. We first illustrate the difference between our \texttt{detrend\_eBIC} and the \texttt{qsreg} using just 2 hours of measurements so that the methods can be easily visualized and the \texttt{qsreg} method can be fit to the entire dataset \Sec{short-app}. We then compare the methods on the complete day shown in \Fig{raw_spod}, fitting the \texttt{qsreg} method to 2 hour non-overlapping windows of the data. We focus on this day because all three sensors were functioning properly. Finally, we examine an entire week of measurements for the two sensors for which data passed was available \Sec{long-app}.
